---
title: "King County Real Estate Exploratory Data Analysis"
author: "Josh Wolfe"
date: "07/08/2021"   
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

```{r echo = FALSE}
# Course: 5210 Data Visualization
# Purpose: Quick Project 1
# Date: July 08, 2021
# Author: Josh Wolfe
```

# Load Libraries

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(gridExtra)
library(scales) 
library(vtable)
library(patchwork)
library(GGally)
library(janitor)
library(choroplethrZip)
library(choroplethr)
```

# Load and Transform Data

```{r}
# Load data
df <- read.csv("qp1_data.csv")

# Transform Numerical variables to factor
df[,'waterfront'] <- factor(df[,'waterfront'])
df[,'view'] <- factor(df[,'view'])
df[,'condition'] <- factor(df[,'condition'])
df[,'yr_built'] <- factor(df[,'yr_built'])
df[,'yr_renovated'] <- factor(df[,'yr_renovated'])
df[,'zip_code'] <- factor(df[,'zip_code'])
df[,'grade'] <- factor(df[,'grade'])

```

# Base EDA Step 1: Uni-variate non-graphical EDA

```{r}
# Look at the top few rows of the data
head(df)
```

+ Data appears to be tidy

    - Each column is of the same variable type and has a unique entry for each row.

```{r}
# Get a breakdown of the variables
str(df)
```

```{r}
# Get a breakdown of the variables
vtable(df)
```

+ Three Numeric variables
+ Three integer variables
+ Seven factor variables

    - Waterfront is a binary variable where 0 is no and 1 is yes
    - View, Grade, and Condition are rankings
    - Yr_built and Yr_renovated are years property was built and renovated
    - Zip Code is the property zip code


```{r}
# Get variable descriptive statistics
summary(df)
```

+ Comments and questions about the data

    - The median property value is $450,000 while the mean is $540,182
    - Property values are positivley skewed likely due to high value outliers
    - Bedrooms are also positivley skewed with a median of 3 but mean of 3.4
    - There is a maximum of 33 bedrooms
    - Interestingly bathrooms are slightly negatively skewed with a median of
    2.25 and mean of 2.1
    - Living space and lot size are also positviely skewed
    - Median/mean floor count is 1.5 with a max of 3.5 floors
    - A majority of properties has no waterfront or view
    - Do properties with a waterfront or view cost more
    - Do properties with views cost more than those with waterfronts
    - Are waterfront properties considered to have a view
    - Most properties are condition 3, which may be average condition
    - Most properties have not be renovated
  
## Potential research questions

+ How are different property values associated with price

    - Square footage of living space or property, floor count, grade,
    condition, etc

+ What does valuation look like from a zip code perspective

+ What does waterfront and view do to valuation

# Base EDA Step 2: Uni-variate graphical EDA

## Values

```{r}
# Create histogram for price
values <- ggplot(data = df, mapping = aes(x = price)) + 
  geom_histogram() +
  labs(title = "Property Values", x = "Price in Dollars",
       y = "Properties") +
  scale_x_continuous(labels = label_number())

values
```
+ Histogram for values show the long right hand side tail

    - shows median is the appropriate statistic to use
    - Skews are due to high value homes
    
## Home Features

```{r}
# Create four graph graphics of bedrooms, bathrooms, living and lot squarefootage
bedrooms <- ggplot(data = df, mapping = aes(x = bedrooms)) + 
  geom_bar() +
  labs(title = "Bedroom Count", x = "bedrooms",
       y = "count")

bathrooms <- ggplot(data = df, mapping = aes(x = bathrooms)) + 
  geom_bar() +
  labs(title = "Bathroom Count", x = "bathrooms",
       y = "count")

sqft_lv <- ggplot(data = df, mapping = aes(x = sqft_living)) + 
  geom_histogram() +
  labs(title = "Living Space", x = "Square Footage",
       y = "count")

sqft_p <- ggplot(data = df, mapping = aes(x = sqft_lot)) + 
  geom_histogram() +
  labs(title = "Lot Square Footage", x = "Square Footage",
       y = "count")

(bathrooms + bedrooms) / (sqft_lv + sqft_p)
```

+ Bar chars show positive skews due to large homes

    - shows median is the appropriate statistic to use for feature values

## Property details

```{r}
# Generate four graph graphic of property details
floors <- ggplot(data = df, mapping = aes(x = floors)) + 
  geom_bar() +
  labs(title = "Floor count", x = "Floors",
       y = "Properties")

condition <- ggplot(data = df, mapping = aes(x = condition)) + 
  geom_bar() +
  labs(title = "Property Condition", x = "Condition",
       y = "properties")

grade <- ggplot(data = df, mapping = aes(x = grade)) + 
  geom_bar() +
  labs(title = "Property Grade", x = "grade",
       y = "properties")

view <- ggplot(data = df, mapping = aes(x = view)) + 
  geom_bar() +
  labs(title = "Property View", x = "View Level",
       y = "Properties")

(floors + view) / (grade + condition)
```

+ Bar charts show that grade is pretty uniform while most properties have
either one or two floors and no view 
+ Most properties have an average condition or greater

## Property Details Continued

```{r}
# Continue previous graphic with remaining variables
waterfront <- ggplot(data = df, mapping = aes(x = waterfront)) + 
  geom_bar() +
  labs(title = "Waterfront Properties", x = "Waterfront",
       y = "Properties")

yr_built <- ggplot(data = df, mapping = aes(x = yr_built)) + 
  geom_bar() +
  labs(title = "Year Property Built", x = "Year",
       y = "properties")

yr_renovated <- ggplot(data = df, mapping = aes(x = yr_renovated)) + 
  geom_bar() +
  labs(title = "Year Property Renovated", x = "Year",
       y = "properties")

zip <- ggplot(data = df, mapping = aes(x = zip_code)) + 
  geom_bar() +
  labs(title = "Property Zip Code", x = "Zip Code",
       y = "Properties")

(waterfront + zip) / (yr_built + yr_renovated)
```

+ Histogram shows that building has increased over the years while most
properties remain renovated. Most properties are not waterfront properties

+ There is no single zip code that stands out in property count.

# Base EDA Step 3: Multi-variate non-graphical EDA

```{r}
# Look at correlational data of quantitative variables
num_corr <- df %>%
  select_if(is.numeric) %>%
  cor() %>%
  round(2) %>%
  kable()

num_corr
```

+ Living square footage has the highest positive correlation with price at 0.7.
This is followed by bathrooms are 0.53

    - Interestingly bathrooms are more positivley correlated with living square
    footage than bedroom count
    - Lot size has very little correlation with price

```{r}
# Examine the overlap of waterfront properties and properties with views
df %>%
  tabyl(view, waterfront) %>%
  adorn_totals(where = c("row", "col")) %>%
  kable()
```
+ Most Waterfront homes have a view ranked 4

    - Do waterfront and view property's values reflect this?

# Base EDA Step 4: Multi-variate graphical EDA

## Categorical Variables

```{r}
# Examine waterfront property information
wv <- df %>%
    ggplot(mapping = aes(x = waterfront, fill = view)) +
    geom_bar(position = "dodge") +
    coord_flip()
  
wc <- df %>%
    ggplot(mapping = aes(x = waterfront, fill = condition)) +
    geom_bar(position = "dodge")
  
wg <- df %>%
    ggplot(mapping = aes(x = waterfront, fill = grade)) +
    geom_bar(position = "dodge")

wv / (wc + wg)

```

+ Waterfront properties are a ver small portion of overall sales

```{r}
# Examine 
vw <- df %>%
    ggplot(mapping = aes(fill = waterfront, y = view)) +
    geom_bar(position = "dodge") +
    coord_flip()
  
vc <- df %>%
    ggplot(mapping = aes(x = view, fill = condition)) +
    geom_bar(position = "dodge")
  
vg <- df %>%
    ggplot(mapping = aes(x = view, fill = grade)) +
    geom_bar(position = "dodge")

vw / (vc + vg)
```

+ Properties with views are a small portion of overall sales

    - Properties with views are usually found with condition 3 and grade 7+

```{r}
# Look at condition and 
grid.arrange(

df %>%
  ggplot(mapping = aes(x = grade, y = price)) +
  geom_point() +
  scale_y_continuous(labels = label_number()),

df %>%
  ggplot(mapping = aes(x = condition, y = price)) +
  geom_point() +
  scale_y_continuous(labels = label_number()),

df %>%
  ggplot(mapping = aes(x = yr_built, y = price)) +
  geom_point() +
  scale_y_continuous(labels = label_number()),

df %>%
  ggplot(mapping = aes(x = zip_code, y = price)) +
  geom_point() +
  scale_y_continuous(labels = label_number())

)
```

+ Grade is also positively correlated while condition interestingly follows
and inverted U, with condition 3 and 4 being high value homes

    - I wonder where the disconnect is for condition 5 homes and why the high
    value homes are not in that catagory
    - High value homes seem to be spread out among multiple zip codes and years
    built

```{r}
df %>%
  group_by(zip_code) %>%
  summarize(median = median(price))
```

+ 98039 (Medina) has the highest median property value ($1,895,000) 
while 98168 (Skyway) and 98002 (Auburn) are tied for lowest ($235,000)

## Quantitative Variables

```{r}

grid.arrange(
  
df %>%
  ggplot(mapping = aes(x = sqft_living, y = price)) +
  geom_point() +
  scale_y_continuous(labels = label_number()),
  
df %>%
  ggplot(mapping = aes(x = sqft_lot, y = price)) +
  geom_point() +
  scale_y_continuous(labels = label_number()),

df %>%
  ggplot(mapping = aes(x = bedrooms, y = price)) +
  geom_point() +
  scale_y_continuous(labels = label_number()),

df %>%
  ggplot(mapping = aes(x = bathrooms, y = price)) +
  geom_point() +
  scale_y_continuous(labels = label_number()),

df %>%
  ggplot(mapping = aes(x = floors, y = price)) +
  geom_point() +
  scale_y_continuous(labels = label_number())

)
```

+ As seen before price is positively correlated with living space while lot size
is not

    - The bedroom outlier has a relatively low selling price which is not
    what I expected
    - Bathrooms and floors follows an expected patters, with more bathrooms being
    associated with a higher selling price and most homes falling into the 2 floor
    range

# Detailed EDA: Questions Raised in Base EDA

## How do waterfront property values compare to non-waterfront?

```{r}
# research question code
# faceted graph

df %>%
  ggplot(mapping = aes(x = price)) +
  geom_boxplot() +
  coord_flip() +
  scale_x_continuous(labels = label_number()) +  
  facet_wrap( ~ waterfront)


```

```{r}
# Get median property values
df %>%
  group_by(waterfront) %>%
  summarize(median = median(price))
```


+ Here we can see that the average values for waterfront properties is higher
than non-waterfront

    - The median price for waterfront properties is $1,400,000 while non-waterfront
    is $450,000

## How does the value of properties with views compare to non-view properties?

```{r}
df %>%
  ggplot(mapping = aes(x = price)) +
  geom_boxplot() +
  coord_flip() +
  scale_x_continuous(labels = label_number()) +  
  facet_wrap( ~ view)
```

```{r}
# Get median property values
df %>%
  group_by(view) %>%
  summarize(median = median(price))
```

+ Property value increases with view with the exception of a view ranked 1 and 2
+ The median value for properties without a view is $432,500 while properties with
a view ranked 5 have a median value of $1,190,000

## What does the overall valuation breakdown of the area look like?

```{r}
final_zip <- df %>%
  group_by(zip_code) %>%
  summarise(zme = median(price)) %>%
  ggplot(aes(x = zip_code, y = zme)) +
  geom_bar(stat = "identity", fill = "orange", width = 0.5) +
  scale_y_continuous(breaks = round(seq(min(df$price), max(df$price), by = 100000),1),
                     labels = scales::dollar_format())+
  geom_text(aes(label=zme), vjust=-0.3, size=3.5)+
  theme_minimal()+
  ggtitle("",
          sub = "") +
  labs(x = "", y = "Property Value")

final_zip
```


# Summary

+ Both waterfront and view are associated with higher property values
+ As view ranking increases property value increases
+ Waterfront and properties with views are a small fraction of overall property sales
+ There is a large difference in median valuation between zip codes

# Create and Save Professional Quality Visual

```{r}
# Waterfront

final_waterfront <- df %>%
  group_by(waterfront) %>%
  summarise(wme = median(price)) %>%
  ggplot(aes(x = waterfront, y = wme)) +
  geom_bar(stat = "identity", fill = "light green", width = 0.5) +
  scale_y_continuous(breaks = round(seq(min(df$price), max(df$price), by = 100000),1),
                     labels = scales::dollar_format())+
  geom_text(aes(label=wme), vjust=-0.3, size=3.5)+
  theme_minimal()+
  ggtitle("",
          sub = "") +
  labs(x = "", y = "Property Value") +
  scale_x_discrete(labels=c("0" = "Non-waterfront", "1" = "Waterfront"))

ggsave(filename = "final_waterfront.png", plot = final_waterfront)
```

```{r}
# View
final_view <- df %>%
  group_by(view) %>%
  summarise(vme = median(price)) %>%
  ggplot(aes(x = view, y = vme)) +
  geom_bar(stat = "identity", fill = "light blue", width = 0.5) +
  scale_y_continuous(breaks = round(seq(min(df$price), max(df$price), by = 100000),1),
                     labels = scales::dollar_format())+
  geom_text(aes(label=vme), vjust=-0.3, size=3.5)+
  theme_minimal()+
  ggtitle("Median Property Values Based on View Rank and Water Frontage",
          sub = "In King County") +
  labs(x = "", y = "Property Value") +
  scale_x_discrete(labels=c("0" = "No View", "1" = "Rank 1", "2" = "Rank 2",
                            
                            "3" = "Rank 3", "4" = "Rank 4"))

final_view_waterfront <- final_view + final_waterfront

ggsave(filename = "final_view_waterfront.png", plot = final_view_waterfront)

final_view_waterfront
```

```{r}
# Create Zip Map
zip_values <- df %>%
  group_by(zip_code) %>%
  summarise(zme = median(price))
colnames(zip_values) <- c("region", "value")
zip_values$region <- as.character(zip_values$region)

kc_fips <- 53033

data(zip.regions)
data(zip_values)
final_zip <- zip_choropleth(zip_values,
               county_zoom = kc_fips,
               title      = "Median King County Property Values by Zip Code",
               legend     = "Values in US Dollars")

ggsave(filename = "final_zip.png", plot = final_zip)

final_zip
```

# Statistical EDA

```{r}
library(Hmisc)
library(MultinomialCI)
library(htmlTable)
```

The steps of inference are:
Gather a representative sample from a population
Establish the null/alternative hypothesis (the research question)
Set confidence level 1−α to balance Type I & II errors
α is the risk level
Test the hypothesis for practical and statistical significance based on the sample
practical significance: How big is the impact or difference of the variable(s)?
statistical significance: How consistent or reliable is the difference?
State the results


## View Rank

+ Null: View has no influence on property value

    - True null: view rank does not impact value
    - False null: view rank does impact sales
    - Type 1: reject true null - ask more for property based on view (inepensive, if expensive use small alpha)
    - Type 2: fail to reject false null - don't consider view in appraisal (expensive, use large alpha)
    
Alternative: View has an influence on property value

```{r}
vchi <- chisq.test(df$view, df$price)
vchi
```

### Practical Significance

```{r}
summary(df$view)
```
There are fewer properties with views than non-view properties. The small sample size may not be practically or significantly significant

### Statistical Significance

The property values are statistically different in relation to view rank with a P value of less than 0.001. Therefore we reject the null hypothesis that property values are not related to view

Non-technical: A properties view ranking has an impact on the property's value
Technical: Property values are statistically different for different view ranks with a P value of less than 0.001

## Waterfrontage

+ Null: Water frontage has no influence on property value
Alternative: Water frontage has an influence on property value

    - True null: water frontage does not impact value
    - False null: waterfrontage does impact sales
    - Type 1: reject true null - ask more for property based on water frontage (inepensive, if expensive use small alpha)
    - Type 2: fail to reject false null - don't consider water frontage in appraisal (expensive, use large alpha)
    
Alternative: Water frontage has an influence on property value

```{r}
wchi <- chisq.test(df$waterfront, df$price)
wchi
```

### Practical Significance

```{r}
summary(df$waterfront)
```
There are only 163 waterfront sales out of 21613 total sales. The sample size is too small to be practically significant

### Statistical Significance

Waterfront property values are statistically different from non-waterfront with a p value of less than 0.001. We reject the null hypothesis that waterfront does not influence property value.


Non-technical: Water frontage has an impact on property value
Technical: Water front property values are statistically different with a P value of less then 0.001




